Best Practices for Single Alchemical Free Energy Calculations [Article v\versionnumber]

### 2025-07-18 paper/manuscript.tex

\newcommand{\versionnumber}{2.0} % you should update the minor version number in preprints and major version number of submissions.
\newcommand{\githubrepository}{\url{https://github.com/alchemistry/alchemical-best-practices}} %this should be the main github repository for this article
\newcommand{\expect}[1]{\left\langle{#1}\right\rangle}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% ARTICLE SETUP
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Best Practices for Single Alchemical Free Energy Calculations [Article v\versionnumber]}
\author[1*]{Antonia S. J. S. Mey}
\author[2]{Bryce K. Allen}
\author[3]{Hannah E. Bruce Macdonald}
\author[3*]{John D. Chodera}
\author[9]{Vytautas Gapsys}
\author[9]{David F. Hahn}
\author[1,10]{Maximilian Kuhn}
\author[1]{Julien Michel}
\author[4*]{David L. Mobley}
\author[5]{Levi N. Naden}
\author[6]{Samarjeet Prasad}
\author[2,7]{Andrea Rizzi}
\author[1]{Jenke Scheen}
\author[11]{David R. Slochower}
\author[8*]{Michael R. Shirts}
\author[9]{Gary Tresadern}
\author[2]{Huafeng Xu}

%
\affil[1]{EaStCHEM School of Chemistry, David Brewster Road, Joseph Black Building, The King's Buildings, Edinburgh, EH9 3FJ, UK}
\affil[2]{Silicon Therapeutics, Boston, MA, USA}
\affil[3]{Computational and Systems Biology Program, Sloan Kettering Institute, Memorial Sloan Kettering Cancer Center, New York NY, USA}
\affil[4]{Departments of Pharmaceutical Sciences and Chemistry, University of California, Irvine, Irvine, USA}
\affil[5]{Molecular Sciences Software Institute, Blacksburg VA, USA}
\affil[6]{National Institutes of Health, Bethesda, MD, USA}
\affil[7]{Tri-Institutional Training Program in Computational Biology and Medicine, New York, NY, USA}
\affil[8]{University of Colorado Boulder, Boulder, CO, USA}
\affil[9]{Computational Chemistry, Janssen Research \& Development, Turnhoutseweg 30, Beerse B-2340, Belgium}
\affil[10]{Cresset, Cambridgeshire, UK}
\affil[11]{Vertex Pharmaceuticals, San Diego, CA, USA}

@@

\orcid{Antonia S. J. S. Mey}{0000-0001-7512-5252}
\orcid{Bryce Allen}{0000-0002-0804-8127}
\orcid{Hannah E. Bruce Macdonald}{0000-0002-5562-6866}
\orcid{John D. Chodera}{0000-0003-0542-119X}
\orcid{Maximilian Kuhn}{0000-0002-2811-3934}
\orcid{Julien Michel}{0000-0003-0360-1760}
\orcid{David L. Mobley}{0000-0002-1083-5533}
\orcid{Levi N. Naden}{0000-0002-3692-5027}
\orcid{Samarjeet Prasad}{0000-0001-8320-6482}
\orcid{Andrea Rizzi}{0000-0001-7693-2013}
\orcid{Jenke Scheen}{0000-0001-9781-0445}
\orcid{David R. Slochower}{0000-0003-3928-5050}
\orcid{Michael R. Shirts}{0000-0003-3249-1097}
\orcid{Gary Tresadern}{0000-0002-4801-1644}
\orcid{Huafeng Xu}{0000-0001-5447-0452}
\orcid{David F. Hahn}{0000-0003-2830-6880}
\orcid{Vytautas Gapsys}{0000-0002-6761-7780}

@@

where $r_{ij}$ is the distance between two particles $i$ and $j$, $\epsilon_{ij}$ and $\sigma_{ij}$ are the Lennard-Jones parameters corresponding to the interaction between particles $i$ and $j$, and $\alpha$ is a constant. In particular, $\alpha=0.5$ is statistically optimal for the specific functional form shown above. This functional form has exactly the property we are looking for: it recovers the Lennard-Jones potential when $\vec{\lambda}=1$, and the at other endpoint ($\vec{\lambda}=0$), it is exactly zero for all $r_{ij}$ everywhere, and as $\vec{\lambda}$ goes to zero, the $\alpha(1-\vec{\lambda})$ term lowers the infinite energy in the core. There are several different variants of the same functional form~\cite{zacharias1994separationshifted, beutler1994avoiding,pham2011identifying}, but the one given in eq.~\ref{eq:softcore} is easy to understand and implement and fairly numerically stable. This functional form is shown in \textbf{C} and \textbf{D} of Fig.~\ref{fig:fig_what_is_lambda}.

@@

Relative calculations introduce additional choices, such as the order in which to modify nonbonded interactions.
A common process in single topology relative calculations is, as noted above, to first remove electrostatic interactions of any atoms which will be deleted, then modify other non-bonded interactions, then restore electrostatic interactions of any atoms which are being inserted. Although this is a simpler path to understand cognitively and can take advantage of the soft-core potential from Eq.~\ref{eq:softcore}, this can lead to more intermediate steps and thus be more computationally expensive.
Other schemes, such as simultaneously changing electrostatic and Lennard-Jones interactions with electrostatic soft-core potentials~\cite{steinbrecher2007nonlinear}, as already discussed above, may be implemented with fewer intermediate states but could require fine-tuning of electrostatic and Lennard-Jones soft-core parameters to avoid numerical instabilities. 
At the time of writing, there has not been conclusive evidence to suggest the separate or simultaneous approach is in general better than the other, all factors considered, so discretion should be left up to the user as to what is viable from both hardware resources, and what the simulation software supports.

@@

\subsubsection{The importance of configurational sampling}
\label{sec:configurational_sampling}
As mentioned previously, free energy differences estimated from alchemical calculations should be independent of the initial system conformation, given sufficient configurational sampling. In cases where the system's slowest degrees of freedom are coupled with the alchemical variable, transitioning between alchemical states can indirectly enhance configurational sampling. This effect is particularly pronounced when certain alchemical states lack the energy barriers present at the fully coupled state (e.g. $\lambda=0$ in Fig. ~\ref{fig:configurational_sampling}\textbf{A}), thereby allowing the system to bypass otherwise prohibitive configurational free energy barriers. For example, in a protein-ligand unbinding process where the kinetic barrier is predominantly associated with the disruption of intermolecular interactions or the separation of binding partners, simulating different intermediate states effectively samples configurations with varying center-of-mass distances, thus mitigating the sampling issue in the configurational space.

Conversely, when the slowest degrees of freedom in the configurational space are largely orthogonal to the alchemical variable, the system can be stuck in its initial metastable state. Such entrapment leads to discrepancies in free energy estimates from simulations initiated from different conformations. This scenario, as illustrated in Fig.~\ref{fig:configurational_sampling}\textbf{B}, is not uncommon especially in flexible systems. For example, it can arise when the free energy difference of interest needs to account for metastable states separated by torsional barriers, such as the solvation free energy of a ligand having different rotamer states, the methylation free energy of a nucleobase in a duplex ~\cite{hsu2023alchemical}, and the binding free energy that involves flexible side chains near the binding site~\cite{la2022alchemical}. Similar sampling challenges have also been reflected in several recent studies, which highlights the sensitivity of alchemical calculations to initial conformations, whether derived from crystallographic structures~\cite{suruzhon2021sensitivity, baumann2024impact} or docking protocols~\cite{cappel2020impact}.

@@

There are multiple strategies to alleviate such sampling challenges in an alchemical calculation. A straightforward but computationally expensive method is to run longer simulations with multiple replicates to increase conformational heterogeneity~\cite{suruzhon2021sensitivity}. Another approach involves performing separate alchemical calculations from different metastable states, then computing the overall free energy difference as the negative log-sum-exp of the Boltzmann weights from each metastable state~\cite{hsu2023alchemical}. However, this method is only applicable to cases like Fig.~\ref{fig:configurational_sampling}\textbf{B}, where the system remains trapped in its initial metastable state throughout the calculation. A generally more effective strategy is to bias relevant configurational collective variables (CVs) during the alchemical simulation. This can be achieved by enhancing sampling within individual alchemical states and swapping conformations between alchemical states (preferably from different metastable states) via replica exchange protocols. More recently, several approaches have been proposed to simultaneously bias both configurational CV and alchemical variables, which can effectively address both scenarios Fig.~\ref{fig:configurational_sampling}\textbf{B} and \textbf{C}. Examples include alchemical metadynamics~\cite{hsu2023alchemical}, $\lambda$-ABF (adaptive biasing force)~\cite{lagardere2024lambda} and its combination~\cite{zhou2025zooming} with well-tempered metadynamics~\cite{barducci2008well}, to name a few.

@@

\subsubsection{Which sampling scheme will work best for my problem?}
\label{sec:sampling_schemes}
Though all alchemical simulations must sample from multiple $\vec{\lambda}$ states, different approaches can be used to achieve this. Fig.~\ref{fig:fig_sampling_scheme} illustrates the four most common schemes. The simplest approach involves running an independent simulation at each of the predefined $\vec{\lambda}$ values (see Fig.~\ref{fig:fig_sampling_scheme}\textbf{A}). This type of scheme is currently used for AMBER TI calculations~\cite{song2019using} and for Sire as implemented in BioSimSpace~\cite{hedges2019biosimspace}. However, if these simulations can be run simultaneously with communication between them, a simple extension allows mixing between these replicas. In this approach, the simulation at each $\vec{\lambda}$ can undergo periodic exchanges with neighboring $\vec{\lambda}$ values. This form of replica exchange, called Hamiltonian replica exchange, is based on ideas developed from Monte Carlo simulations of spin glasses by Swendsen and Wang~\cite{swendsen1986replica}. With the Metropolis-Hastings acceptance criterion for exchanges, the generated ensemble of all replicas still samples from the Boltzmann distribution for each replica. This approach has been used in many different contexts for molecular simulations~\cite{sugita2000multidimensional,sugita1999replicaexchange, woods2003development, jiang2010free}. The basic idea of the replica exchange scheme is shown in Fig.~\ref{fig:fig_sampling_scheme}\textbf{B}. It is supported in various software packages that provide alchemical implementations, such as GROMACS~\cite{aldeghi2015accurate}, GROMOS~\cite{hritz2008hamiltonian,hritz2007optimization}, FEP+~\cite{wang2015accurate}, and NAMD~\cite{jiang2019computing}.

@@

The non-equilibrium switching (Fig.~\ref{fig:fig_sampling_scheme}\textbf{D}) approach makes use of short out-of-equilibrium simulations~\cite{aldeghi2018accurate}. In this protocol, only end state $\vec{\lambda}$ replicas ($\vec{\lambda}$=0, $\vec{\lambda}=1$) are simulated at equilibrium; intermediate information is generated from non-equilibrium simulations that rapidly transition between end states. 
Over the course of each transition, Hamiltonian derivative with respect to $\lambda$ is recorded; subsequently the curves are integrated to yield work distributions for the forward and reverse processes. Crooks Fluctuation Theorem~\cite{crooks1999entropy} allows estimating free energy difference between the end states from the generated work distributions.~\cite{shirts2003equilibrium}

The non-equilibrium (NEQ) approaches offer some advantages over the equilibrium sampling based methods. For example, the NEQ protocol allows equilibrium simulations of each end-state to be performed independently without pre-defining ligand network: the transitions between the ligands can be generated afterwards. This eliminates the redundant equilibrium sampling, e.g. in a starmap network, FEP based approaches would require sampling the central ligand for each edge, while NEQ would sample it once and connect to the other ligands afterwards. The NEQ protocol also provides flexibility in combining different structures in free energy calculations, e.g. simulations in states at $\lambda=0$ and $\lambda=1$ may be initialized with the independently generated ligand coordinates. Another example would be using true apo and holo protein structures as the end states to initialize simulations for absolute binding free energy calculations.

@@

Currently, we recommend using Hamiltonian replica exchange type sampling schemes (Fig.~\ref{fig:fig_sampling_scheme} \textbf{B}). If these are not available in the code of choice, running independent simulations at different $\vec{\lambda}$ values can be acceptable, especially when configurational sampling is fast (Fig.~\ref{fig:fig_sampling_scheme} \textbf{A}). Single replica schemes and non-equilibrium schemes are not as established yet because of potential failure modes, but are very promising for use in the near future.

\subsubsection{How long should I run my simulation for and what information should be saved?}
\label{sec:sim_length_information_kept}
Before launching alchemical free energy calculations it is wise to consider how convergence and completion will be assessed. Different conditions on when to stop alchemical free energy calculations should be determined, and this may require several iterative checks and therefore modifications to the calculation protocol.
One useful metric to use for termination is the expected or desired uncertainty of a desired free energy estimate, though care must be exercised should the uncertainty estimate prove unreliable.
In particular, if the rate of change in the free energy estimate is significant when this condition is met, the simulation may not be locally converged, and more sampling may be necessary to determine a stable free energy estimate which is no longer changing significantly over time.

@@

The condition of minimizing the statistical uncertainty of different free energy estimators below a sufficient threshold should be one metric monitored over the simulation. This can be done through the uncertainty estimator built into certain analysis tools such as MBAR, or through more general statistical tools like bootstrap sampling. It should be noted however, that uncertainty estimates have the same limitations as other metrics of convergence, as they are only an uncertainty based on the phase space sampled so far in a simulation, and cannot account for states not sampled, and it is worth considering that they will be an underestimation of the true uncertainty.

A target statistical uncertainty should be chosen at the onset of the simulation to avoid excessively long simulations, or falling into the trap of running until the free energy estimate is "good enough," which is subjective and has no defined criteria. This could be a fixed value such as $0.20$ kcal/mol, or a functional quantity such as "below $0.5$ kcal/mol and $10\%$ of the free energy estimate." The user does not need to monitor this information in real-time and can choose to run simulations for fixed duration (either time or number of samples) and run analysis on the data collected thus far. If more samples are needed, the simulations can be resumed, or, started again in different initial conditions.

@@

The uncertainty in the free energy, for example, can be estimated in multiple ways, e.g. through standard error propagation methods (including MBAR's estimator, which is based on the same principles as standard error propagation), through bootstrap methods, and through multiple independent runs. 
Independent of how the property is estimated, it is important to remember that results of any free energy analysis are \textit{estimations of the given property}, not the true underlying value of the property itself. 
These estimators are usually consistent estimators, meaning they will converge to the true answer in the limit of sufficient sampling, not necessarily unbiased ones though.
As such, it is a good idea to subject different estimators to the same data to see if they yield either the same estimate (within error and bias), or if they fluctuate wildly. See, for example, the potential of mean force with respect to $\vec{\lambda}$ estimated from a bound simulation of a Tyk2 ligand pair of Wang et al.~\cite{wang2015accurate} for both the MBAR and TI estimators, as seen in Fig.~\ref{fig:pmf}.
This is not a perfect method as some estimators, such as exponential averaging, will converge significantly more slowly, relative to more accurate estimators like MBAR. 
Therefore, it is a good idea to apply the estimators to different fractions of the data to see if the main estimator of free energy you have chosen is stable.

@@

>>>>

\newcommand{\versionnumber}{2.0} % you should update the minor version number in preprints and major version number of submissions.
\newcommand{\githubrepository}{\url{https://github.com/alchemistry/alchemical-best-practices}} %this should be the main github repository for this article
\newcommand{\expect}[1]{\left\langle{#1}\right\rangle}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% ARTICLE SETUP
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Best Practices for Single Alchemical Free Energy Calculations [Article v\versionnumber]}
\author[1*]{Antonia S. J. S. Mey}
\author[2]{Bryce K. Allen}
\author[11]{Agastya P. Bhati}
\author[3]{Hannah E. Bruce Macdonald}
\author[3*]{John D. Chodera}
\author[11, 12, 13]{Peter V. Coveney}
\author[9]{Vytautas Gapsys}
\author[9]{David F. Hahn}
\author[1,10]{Maximilian Kuhn}
\author[1]{Julien Michel}
\author[4*]{David L. Mobley}
\author[5]{Levi N. Naden}
\author[6]{Samarjeet Prasad}
\author[2,7]{Andrea Rizzi}
\author[1]{Jenke Scheen}
\author[14]{David R. Slochower}
\author[8*]{Michael R. Shirts}
\author[9]{Gary Tresadern}
\author[2]{Huafeng Xu}

%
\affil[1]{EaStCHEM School of Chemistry, David Brewster Road, Joseph Black Building, The King's Buildings, Edinburgh, EH9 3FJ, UK}
\affil[2]{Silicon Therapeutics, Boston, MA, USA}
\affil[3]{Computational and Systems Biology Program, Sloan Kettering Institute, Memorial Sloan Kettering Cancer Center, New York NY, USA}
\affil[4]{Departments of Pharmaceutical Sciences and Chemistry, University of California, Irvine, Irvine, USA}
\affil[5]{Molecular Sciences Software Institute, Blacksburg VA, USA}
\affil[6]{National Institutes of Health, Bethesda, MD, USA}
\affil[7]{Tri-Institutional Training Program in Computational Biology and Medicine, New York, NY, USA}
\affil[8]{University of Colorado Boulder, Boulder, CO, USA}
\affil[9]{Computational Chemistry, Janssen Research \& Development, Turnhoutseweg 30, Beerse B-2340, Belgium}
\affil[10]{Cresset, Cambridgeshire, UK}
\affil[11]{Centre for Computational Sciences, Department of Chemistry, University College London, London WC1H 0AJ, United Kingdom}
\affil[12]{Advanced Research Computing Centre, University College London, London WC1H 9BT, United Kingdom}
\affil[13]{Computational Science Laboratory, Institute for Informatics, Faculty of Science, University of Amsterdam, Amsterdam 1012, The Netherlands}
\affil[14]{Vertex Pharmaceuticals, San Diego, CA, USA}

@@

\orcid{Antonia S. J. S. Mey}{0000-0001-7512-5252}
\orcid{Bryce Allen}{0000-0002-0804-8127}
\orcid{Hannah E. Bruce Macdonald}{0000-0002-5562-6866}
\orcid{John D. Chodera}{0000-0003-0542-119X}
\orcid{Maximilian Kuhn}{0000-0002-2811-3934}
\orcid{Julien Michel}{0000-0003-0360-1760}
\orcid{David L. Mobley}{0000-0002-1083-5533}
\orcid{Levi N. Naden}{0000-0002-3692-5027}
\orcid{Samarjeet Prasad}{0000-0001-8320-6482}
\orcid{Andrea Rizzi}{0000-0001-7693-2013}
\orcid{Jenke Scheen}{0000-0001-9781-0445}
\orcid{David R. Slochower}{0000-0003-3928-5050}
\orcid{Michael R. Shirts}{0000-0003-3249-1097}
\orcid{Gary Tresadern}{0000-0002-4801-1644}
\orcid{Huafeng Xu}{0000-0001-5447-0452}
\orcid{David F. Hahn}{0000-0003-2830-6880}
\orcid{Vytautas Gapsys}{0000-0002-6761-7780}
\orcid{Agastya P. Bhati}{0000-0003-4539-4819}
\orcid{Peter V. Coveney}{0000-0002-8787-7256}

@@

where $r_{ij}$ is the distance between two particles $i$ and $j$, $\epsilon_{ij}$ and $\sigma_{ij}$ are the Lennard-Jones parameters corresponding to the interaction between particles $i$ and $j$, and $\alpha$ is a constant. In particular, $\alpha=0.5$ is statistically optimal for the specific functional form shown above. This functional form has exactly the property we are looking for: it recovers the Lennard-Jones potential when $\vec{\lambda}=1$, and the at other endpoint ($\vec{\lambda}=0$), it is exactly zero for all $r_{ij}$ everywhere, and as $\vec{\lambda}$ goes to zero, the $\alpha(1-\vec{\lambda})$ term lowers the infinite energy in the core. There are several different variants of the same functional form~\cite{zacharias1994separationshifted, beutler1994avoiding,pham2011identifying,Lee_JChemTheoryComput_2020_v16_p5512,Tsai_JChemTheoryComput_2023_v19_p640}, but the one given in eq.~\ref{eq:softcore} is easy to understand and implement and fairly numerically stable. This functional form is shown in \textbf{C} and \textbf{D} of Fig.~\ref{fig:fig_what_is_lambda}.

@@

Relative calculations introduce additional choices, such as the order in which to modify nonbonded interactions.
A common process in single topology relative calculations is, as noted above, to first remove electrostatic interactions of any atoms which will be deleted, then modify other non-bonded interactions, then restore electrostatic interactions of any atoms which are being inserted. Although this is a simpler path to understand cognitively and can take advantage of the soft-core potential from Eq.~\ref{eq:softcore}, this can lead to more intermediate steps and thus be more computationally expensive.
Other schemes, such as simultaneously changing electrostatic and Lennard-Jones interactions with electrostatic soft-core potentials~\cite{steinbrecher2007nonlinear}, as already discussed above, may be implemented with fewer intermediate states but could require fine-tuning of electrostatic and Lennard-Jones soft-core parameters to avoid numerical instabilities~\cite{Tsai_JChemTheoryComput_2023_v19_p640}. 
At the time of writing, there has not been conclusive evidence to suggest the separate or simultaneous approach is in general better than the other, all factors considered, so discretion should be left up to the user as to what is viable from both hardware resources, and what the simulation software supports.

@@

\subsubsection{The importance of configurational sampling}
\label{sec:configurational_sampling}
As mentioned previously, free energy differences estimated from alchemical calculations should be independent of the initial system conformation, given sufficient configurational sampling. In cases where the system's slowest degrees of freedom are coupled with the alchemical variable, transitioning between alchemical states can indirectly enhance configurational sampling. This effect is particularly pronounced when certain alchemical states, such as the decoupled state or loosely coupled states, lack the energy barriers often present at the fully coupled state (e.g. $\lambda=0$ in Fig. ~\ref{fig:configurational_sampling}\textbf{A}), thereby allowing the system to bypass otherwise prohibitive configurational free energy barriers. For example, in a protein-ligand unbinding process where the kinetic barrier is predominantly associated with the disruption of intermolecular interactions or the separation of binding partners, simulating different intermediate states effectively samples configurations with varying center-of-mass distances, thus mitigating the sampling issue in the configurational space.

Conversely, when the slowest degrees of freedom in the configurational space are largely orthogonal to the alchemical variable, the system can be stuck in its initial metastable state. Such entrapment leads to discrepancies in free energy estimates from simulations initiated from different conformations. This scenario, as illustrated in Fig.~\ref{fig:configurational_sampling}\textbf{B}, is not uncommon especially in flexible systems. For example, it can arise when the free energy difference of interest needs to account for metastable states separated by torsional barriers, such as the solvation free energy of a ligand having different rotamer states, the methylation free energy of a nucleobase in a duplex ~\cite{hsu2023alchemical}, and the binding free energy that involves flexible side chains near the binding site~\cite{la2022alchemical}. Similar sampling challenges have also been reflected in several recent studies, which highlight the sensitivity of alchemical calculations to initial conformations, whether derived from crystallographic structures~\cite{suruzhon2021sensitivity, baumann2024impact} or docking protocols~\cite{cappel2020impact}.

@@

There are multiple strategies to alleviate such sampling challenges in an alchemical calculation. A straightforward but computationally expensive method is to run longer simulations with multiple replicates to increase conformational heterogeneity~\cite{suruzhon2021sensitivity}. Another approach, known as integration over parts,\cite{jayachandran2006parallelized}, involves performing separate alchemical calculations limited to a series of binding macrostates and  then computing the overall free energy difference using the free energy combination formula~\cite{gallicchio2011recentadv,hsu2023alchemical}, i.e., computing the negative log-sum-exp of the Boltzmann weights from each macrostate. In the integration over parts approach, restraining potentials can be used to restrict the sampling to specific macrostates,\cite{khuttan2023taming}. However, in practice, restraining potentials are sometimes omitted when the states are known to be separated by large free energy barriers and do not interconvert within the simulation time scale, such as cases like Fig.~\ref{fig:configurational_sampling}\textbf{B}. A generally more effective strategy is to bias relevant configurational collective variables (CVs) during the alchemical simulation. This can be achieved by enhancing sampling within individual alchemical states and swapping conformations between alchemical states (preferably from states separated by large free energy barriers) via replica exchange protocols. More recently, several approaches have been proposed to simultaneously bias both configurational CV and alchemical variables, which can effectively address both scenarios Fig.~\ref{fig:configurational_sampling}\textbf{B} and \textbf{C}. Examples include alchemical metadynamics~\cite{hsu2023alchemical}, $\lambda$-ABF (adaptive biasing force)~\cite{lagardere2024lambda}, and optimized alchemical enhanced sampling (ACES)~\cite{Lee_JChemTheoryComput_2023_v19_p472}  and approaches that integrate well-tempered metadynamics~\cite{barducci2008well} with alchemical methods~\cite{zhou2025zooming, khuttan2024make}, to name a few.

@@

\subsubsection{Which $\lambda$-sampling scheme will work best for my problem?}
\label{sec:sampling_schemes}
Though all alchemical simulations must sample from multiple $\vec{\lambda}$ states, different approaches can be used to achieve this. Fig.~\ref{fig:fig_sampling_scheme} illustrates the four most common schemes. The simplest approach involves running an independent simulation at each of the predefined $\vec{\lambda}$ values (see Fig.~\ref{fig:fig_sampling_scheme}\textbf{A}). This type of scheme is currently used for some AMBER TI calculations~\cite{song2019using} and for Sire as implemented in BioSimSpace~\cite{hedges2019biosimspace}. However, if these simulations can be run simultaneously with communication between them, a simple extension allows mixing between these replicas. In this approach, the simulation at each $\vec{\lambda}$ can undergo periodic exchanges with neighboring $\vec{\lambda}$ values. This form of replica exchange, called Hamiltonian replica exchange, is based on ideas developed from Monte Carlo simulations of spin glasses by Swendsen and Wang~\cite{swendsen1986replica}. With the Metropolis-Hastings acceptance criterion for exchanges, the generated ensemble of all replicas still samples from the Boltzmann distribution for each replica. This approach has been used in many different contexts for molecular simulations~\cite{sugita2000multidimensional,sugita1999replicaexchange, woods2003development, jiang2010free}. The basic idea of the replica exchange scheme is shown in Fig.~\ref{fig:fig_sampling_scheme}\textbf{B}. It is supported in various software packages that provide alchemical implementations, such as GROMACS~\cite{aldeghi2015accurate}, GROMOS~\cite{hritz2008hamiltonian,hritz2007optimization},
AMBER~\cite{Lee_JChemInfModel_2020_v60_p5595}, FEP+~\cite{wang2015accurate}, and NAMD~\cite{jiang2019computing}.
A specific variant of Hamiltonian replica exchange is FEP/REST2~\cite{rest2, feprest} that involves ``heating'' a small portion of the protein-ligand system to enhance the sampling within the binding pocket. It has been promoted as a great way to enhance sampling and is employed by default in Schr\"{o}dinger's FEP+ package. However, it should be noted that several studies have questioned its reliability and reported that it degrades the quality of free energy predictions as compared to those based on standard MD simulations without any enhanced sampling~\cite{wan2020fep+, bhati2022}. Therefore, a blind application of this sampling scheme is not recommended and it must be used with caution if at all. A variant of this method that overcomes some of the reported issues with FEP/REST2 is called alchemical enhanced sampling (ACES)~\cite{Lee_JChemTheoryComput_2023_v19_p472}. ACES creates localized targeted "enhanced sampling states" that are connected to the real state end-points through HREMD.  The method leverages the dual topology framework to produce a concerted “counter-diffusion” of alchemical states that leads to minimal rearrangement of the environment along the alchemical path~\cite{York_ACSPhysChemAu_2023_v3_p478}. The computational overhead of the method is negligible relative to more conventional alchemical free energy simulations with HREMD.

@@

The non-equilibrium (NEQ) switching (Fig.~\ref{fig:fig_sampling_scheme}\textbf{D}) approach makes use of short out-of-equilibrium simulations~\cite{aldeghi2018accurate}. In this protocol, only end state $\vec{\lambda}$ replicas ($\vec{\lambda}$=0, $\vec{\lambda}=1$) are simulated at equilibrium; intermediate information is generated from non-equilibrium simulations that rapidly transition between end states. 
Over the course of each transition, Hamiltonian derivative with respect to $\lambda$ is recorded; subsequently the curves are integrated to yield work distributions for the forward and reverse processes. Crooks Fluctuation Theorem~\cite{crooks1999entropy} allows estimating free energy difference between the end states from the generated work distributions.~\cite{shirts2003equilibrium}

The NEQ approaches offer some advantages over the equilibrium sampling based methods. For example, the NEQ protocol allows equilibrium simulations of each end-state to be performed independently without pre-defining a ligand network: the transitions between the ligands can be generated afterwards. This eliminates the redundant equilibrium sampling if one end state is present in multiple perturbations. Equilibrium sampling based approaches would require sampling the central ligand for each edge, while NEQ would sample it once and connect to the other ligands afterwards. The NEQ protocol also provides a simple approach in combining different structures in free energy calculations, e.g. simulations in states at $\lambda=0$ and $\lambda=1$ may be initialized with the independently generated ligand coordinates. Another example would be using true apo and holo protein structures as the end states to initialize simulations for absolute binding free energy calculations. 
Bhati \textit{et al.} recently published two studies directly comparing the equilibrium (EQ) and NEQ approaches at unprecedented scales for both RBFE and ABFE calculations using their ensemble simulation-based TIES method~\cite{wan2023eqvsneq, bhati2025}. Further, they systematically studied the effect of various parameters involved in setting up calculations for both approaches and recommended optimal protocols in each case for industrial applications. They highlighted several limitations of the NEQ approach that should be accounted for while performing such calculations to ensure reliability of predictions.

@@

Currently, we recommend using Hamiltonian replica exchange type sampling schemes (Fig.~\ref{fig:fig_sampling_scheme} \textbf{B}). If these are not available in the code of choice, running independent simulations at different $\vec{\lambda}$ values can be acceptable, especially when configurational sampling is fast (Fig.~\ref{fig:fig_sampling_scheme} \textbf{A}). Single replica schemes and non-equilibrium schemes are not as established yet because of potential failure modes, but are very promising for use in the near future. 
Further, ensemble simulations are recommended irrespective of the choice of sampling scheme to ensure reliability of results~\cite{bhati2018, bhati2019, wan2020fep+, wade2022, wan2021uq, bhati2025, wan2023eqvsneq}.

\subsubsection{How long should I run my simulation for and what information should be saved?}
\label{sec:sim_length_information_kept}
Before launching alchemical free energy calculations it is wise to consider how convergence and completion will be assessed. Different conditions on when to stop alchemical free energy calculations should be determined, and this may require several iterative checks and therefore modifications to the calculation protocol.
One useful metric to use for termination is the expected or desired uncertainty of a desired free energy estimate, though care must be exercised should the uncertainty estimate prove unreliable.
In particular, if the rate of change in the free energy estimate is significant when this condition is met, the simulation may not be locally converged, and more sampling may be necessary to determine a stable free energy estimate which is no longer changing significantly over time. 
A few studies reported systematic variation of simulation lengths using ensemble simulations over an extended range to provide recommendations on appropriate values applicable for most protein-ligand systems. The authors suggest using 4 ns and 10 ns for RBFE~\cite{bhati2017, bhati2022, wan2023eqvsneq} and ABFE~\cite{bhati2025} calculations respectively. These values may be used as a general guideline.

@@

The condition of minimizing the statistical uncertainty of different free energy estimators below a sufficient threshold should be one metric monitored over the simulation. This can be done through the uncertainty estimator built into certain analysis tools such as MBAR, or through more general statistical tools like bootstrap sampling. It should be noted however, that uncertainty estimates have the same limitations as other metrics of convergence, as they are only an uncertainty based on the phase space sampled so far in a simulation, and cannot account for states not sampled, and it is worth considering that they will be an underestimation of the true uncertainty~\cite{wan2020fep+, wan2021uq, vassaux2021, gapsys2020large, wade2022, bhati2017, bhati2022, bhati2025, wan2023eqvsneq, bhati2018, bhati2019}.
A target statistical uncertainty should be chosen at the onset of the simulation to avoid excessively long simulations, or falling into the trap of running until the free energy estimate is "good enough," which is subjective and has no defined criteria. This could be a fixed value such as $0.20$ kcal/mol, or a functional quantity such as "below $0.5$ kcal/mol and $10\%$ of the free energy estimate." The user does not need to monitor this information in real-time and can choose to run simulations for fixed duration (either time or number of samples) and run analysis on the data collected thus far. If more samples are needed, the simulations can be resumed, or, started again in different initial conditions.
Bhati \textit{et al.} recommended such a scheme for large scale RBFE and ABFE calculations using TIES~\cite{bhati2022, bhati2025, wan2023eqvsneq, coveney2022corners}. They suggest a set of optimal parameters (ensemble size, simulation length, number of lambda-states, number and length of NEQ transitions, etc.) to be used initially. Then, only those systems which do not satisfy pre-defined criteria (uncertainties below 0.5 kcal/mol and overlap between forward and reverse work values in case of NEQ calculations) should be further evaluated using additional calculations (in the way suggested) until these criteria are met. Such a scheme optimizes the overall computational efficiency for the full dataset in large scale applications.

@@

The uncertainty in the free energy, for example, can be estimated in multiple ways, e.g. through standard error propagation methods (including MBAR's estimator, which is based on the same principles as standard error propagation), through bootstrap methods, and through ensemble simulations. However, as we have repeatedly pointed out, only ensemble simulations can capture the true uncertainty in MD based predictions. All other methods underestimate the true uncertainty as reported by several authors~\cite{wan2020fep+, wan2021uq, vassaux2021, gapsys2020large, wade2022, bhati2017, bhati2022, bhati2025, wan2023eqvsneq, bhati2018, bhati2019}. 
Independent of how the property is estimated, it is important to remember that results of any free energy analysis are \textit{estimations of the given property}, not the true underlying value of the property itself. 
These estimators are usually consistent estimators, meaning they will converge to the true answer in the limit of sufficient sampling, not necessarily unbiased ones though.
As such, it is a good idea to subject different estimators to the same data to see if they yield either the same estimate (within error and bias), or if they fluctuate wildly. See, for example, the potential of mean force with respect to $\vec{\lambda}$ estimated from a bound simulation of a Tyk2 ligand pair of Wang et al.~\cite{wang2015accurate} for both the MBAR and TI estimators, as seen in Fig.~\ref{fig:pmf}.
This is not a perfect method as some estimators, such as exponential averaging, will converge significantly more slowly, relative to more accurate estimators like MBAR. 
Therefore, it is a good idea to apply the estimators to different fractions of the data to see if the main estimator of free energy you have chosen is stable.

@@

\subsection{Alchemical Transfer Method for Absolute and Relative Binding Free Energy Calculations}

\subsubsection{Introduction}

The Alchemical Transfer Method (ATM)~\cite{wu2021alchemical,azimi2022relative} is an alchemical framework for estimating binding free energies that targets generality with a rigorous and straightforward implementation, with as few additions as possible to standard non-alchemical models of molecular systems. Additionally, the method targets complex ligand transformations in drug design, involving scaffold-hopping and charge-changing transformations that are not well supported by traditional algorithms. ATM's overarching philosophy is to deploy robust and rigorous free energy estimation algorithms with the simplest possible procedures for developers and end users. One critical design principle is the seamless support for advanced molecular potential energy functions, such as quantum-mechanical, polarizable, and neural network potentials, in free energy-based drug discovery applications.
